---
title: "Getting started with rjsf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rjsf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

`rjsf` provides an R interface to the Python `jsf` package for Jump-Switch-Flow simulation.

The main function is:

```{r}
jsf_simulate()
```

This function lets users define a compartmental reaction model in R, pass it to the Python `jsf` backend through `reticulate`, and return the simulation output as either:

- a regular R `data.frame`
- a structured `JSFResult` object

A `JSFResult` object supports:

```{r}
print(result)
summary(result)
plot(result)
```

# Checking Python `jsf`

Before running simulations, check whether the Python `jsf` package is available in the current `reticulate` environment.

```{r}
library(rjsf)

jsf_available()
```

If this returns `FALSE`, check that your Python environment contains the Python `jsf` package and that `reticulate` is pointing to the correct Python environment.

# Lotka-Volterra predator-prey model

The Lotka-Volterra model describes interaction between two species:

- `prey`
- `predator`

The reactions are:

1. Prey birth: `Prey -> 2 Prey`
2. Predator death: `Predator -> empty`
3. Predation: `Prey + Predator -> 2 Predator`

The state vector is:

```{r}
x = c(prey, predator)
```

## Define the rate function

The Python `jsf` backend expects the rate function to be a Python callable. We create this using `reticulate::py_eval()`.

```{r}
library(reticulate)

rates_lv <- reticulate::py_eval(
  "lambda x, t: [2.0 * x[0], 1.5 * x[1], 0.05 * x[0] * x[1]]"
)
```

The rates correspond to:

```text
2.0 * prey                # prey birth
1.5 * predator            # predator death
0.05 * prey * predator    # predation
```

## Define the stoichiometry

The reactant matrix describes what each reaction consumes.

```{r}
reactant_lv <- matrix(
  c(
    1, 0,
    0, 1,
    1, 1
  ),
  ncol = 2,
  byrow = TRUE
)

reactant_lv
```

The product matrix describes what each reaction produces.

```{r}
product_lv <- matrix(
  c(
    2, 0,
    0, 0,
    0, 2
  ),
  ncol = 2,
  byrow = TRUE
)

product_lv
```

Internally, `jsf_simulate()` computes the net stoichiometry matrix:

```text
nu = product - reactant
```

For this model, the net changes are:

```text
Prey -> 2 Prey              [+1,  0]
Predator -> empty           [ 0, -1]
Prey + Predator -> 2 Predator [-1, +1]
```

## Run the simulation

```{r}
result_lv <- jsf_simulate(
  x0 = c(prey = 50, predator = 10),
  rates = rates_lv,
  reactant = reactant_lv,
  product = product_lv,
  do_disc = c(1, 1),
  t_max = 10,
  dt = 0.01,
  switching_threshold = c(30, 30),
  species_names = c("prey", "predator"),
  return_type = "JSFResult"
)
```

Here:

```text
x0 = c(prey = 50, predator = 10)
```

sets the initial state.

```text
do_disc = c(1, 1)
```

means both species may be treated discretely or stochastically when needed.

```text
switching_threshold = c(30, 30)
```

sets the switching threshold for both species.

## Inspect the result

```{r}
result_lv
```

```{r}
summary(result_lv)
```

```{r}
plot(result_lv)
```

# SIR epidemic model

The SIR model describes the spread of an infectious disease through three compartments:

- `S`: susceptible
- `I`: infectious
- `R`: recovered

The reactions are:

1. Infection: `S + I -> 2I`
2. Recovery: `I -> R`

The state vector is:

```{r}
x = c(S, I, R)
```

## Define model parameters

```{r}
N <- 1000
beta <- 0.5
gamma <- 0.1
```

where:

```text
N      = total population
beta   = infection rate
gamma  = recovery rate
```

## Define the rate function

```{r}
rates_sir <- reticulate::py_eval(sprintf(
  "lambda x, t: [%f * x[0] * x[1] / %f, %f * x[1]]",
  beta,
  N,
  gamma
))
```

The rates correspond to:

```text
beta * S * I / N    # infection
gamma * I           # recovery
```

## Define the stoichiometry

```{r}
reactant_sir <- matrix(
  c(
    1, 1, 0,
    0, 1, 0
  ),
  ncol = 3,
  byrow = TRUE
)

reactant_sir
```

```{r}
product_sir <- matrix(
  c(
    0, 2, 0,
    0, 0, 1
  ),
  ncol = 3,
  byrow = TRUE
)

product_sir
```

The net changes are:

```text
S + I -> 2I    [-1, +1, 0]
I -> R         [ 0, -1, +1]
```

## Run the simulation

```{r}
result_sir <- jsf_simulate(
  x0 = c(S = N - 10, I = 10, R = 0),
  rates = rates_sir,
  reactant = reactant_sir,
  product = product_sir,
  do_disc = c(0, 1, 0),
  t_max = 60,
  dt = 0.1,
  switching_threshold = c(N, 10, N),
  species_names = c("S", "I", "R"),
  return_type = "JSFResult"
)
```

Here:

```text
do_disc = c(0, 1, 0)
```

means the infectious compartment may be treated discretely or stochastically.

This is useful because stochastic effects are often most important when the number of infectious individuals is small.

## Inspect the result

```{r}
result_sir
```

```{r}
summary(result_sir)
```

```{r}
plot(result_sir)
```

# Returning a data frame instead

By default, `jsf_simulate()` can return a regular data frame.

```{r}
df_lv <- jsf_simulate(
  x0 = c(prey = 50, predator = 10),
  rates = rates_lv,
  reactant = reactant_lv,
  product = product_lv,
  do_disc = c(1, 1),
  t_max = 10,
  dt = 0.01,
  switching_threshold = c(30, 30),
  species_names = c("prey", "predator"),
  return_type = "data.frame"
)

head(df_lv)
```

A data frame can also be converted into a `JSFResult` object later.

```{r}
result_from_df <- as_jsf_result(
  df_lv,
  method = "operator-splitting",
  config = list(
    model = "Lotka-Volterra",
    dt = 0.01,
    switching_threshold = c(30, 30)
  )
)

result_from_df
```

# Summary

The main workflow is:

```text
Define initial state
  -> define rate function
  -> define reactant and product matrices
  -> call jsf_simulate()
  -> return data.frame or JSFResult
  -> use print(), summary(), and plot()
```

The main user-facing function is:

```{r}
jsf_simulate()
```

The `JSFResult` output is useful when users want a structured object that stores both simulation trajectories and metadata.
