---
title: "Getting started with rjsf"
description: "Set up Python JSF, define compartmental models, and run Jump-Switch-Flow simulations from R."
toc: true
toc-depth: 3
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

## Introduction

`rjsf` provides an R interface to the Python `jsf` package for Jump-Switch-Flow simulation.

The main user-facing function is `jsf_simulate()`. It lets users:

1. Define a compartmental reaction model in R.
2. Pass the model to the Python `jsf` backend through `reticulate`.
3. Return the simulation output as either:

   * a regular R `data.frame`; or
   * a structured `JSFResult` object.

A `JSFResult` object supports standard methods such as:

```{r}
print(result)
summary(result)
plot(result)
```

## Set up Python `jsf`

`rjsf` uses `reticulate` to communicate with Python. The Python `jsf` package must therefore be installed in the same Python environment used by `reticulate`.

One option is to install the dependencies from R:

```{r}
install.packages("reticulate")

reticulate::virtualenv_create("rjsf-env")

reticulate::py_install(
  packages = c(
    "numpy",
    "python-libsbml",
    "ipykernel"
  ),
  envname = "rjsf-env",
  method = "virtualenv"
)

reticulate::py_install(
  packages = "git+https://github.com/DGermano8/jsf.git",
  envname = "rjsf-env",
  method = "virtualenv",
  pip = TRUE
)

reticulate::use_virtualenv(
  "rjsf-env",
  required = TRUE
)
```

You can inspect the Python environment selected by `reticulate` with:

```{r}
reticulate::py_config()
```

## Check Python `jsf`

Load `rjsf` and check whether the Python `jsf` module is available:

```{r}
library(rjsf)

jsf_available()
```

A correctly configured environment should return:

```text
[1] TRUE
```

If it returns `FALSE`, check that:

* Python `jsf` is installed;
* it is installed in the environment currently used by `reticulate`; and
* `reticulate::py_config()` reports the expected Python executable.

## Lotka–Volterra predator–prey model

The Lotka–Volterra model describes interactions between two species:

* `prey`;
* `predator`.

The model contains three reactions:

1. Prey birth: `Prey -> 2 Prey`
2. Predator death: `Predator -> ∅`
3. Predation: `Prey + Predator -> 2 Predator`

The state vector is:

```{r}
x <- c(prey, predator)
```

### Define the rate function

The Python `jsf` backend expects the rate function to be a Python callable. We create one using `reticulate::py_eval()`:

```{r}
rates_lv <- reticulate::py_eval(
  "lambda x, t: [2.0 * x[0], 1.5 * x[1], 0.05 * x[0] * x[1]]"
)
```

The three rates correspond to:

```text
2.0 * prey                # prey birth
1.5 * predator            # predator death
0.05 * prey * predator    # predation
```

### Define the stoichiometry

The reactant matrix describes what each reaction consumes:

```{r}
reactant_lv <- matrix(
  c(
    1, 0,
    0, 1,
    1, 1
  ),
  ncol = 2,
  byrow = TRUE
)

reactant_lv
```

The product matrix describes what each reaction produces:

```{r}
product_lv <- matrix(
  c(
    2, 0,
    0, 0,
    0, 2
  ),
  ncol = 2,
  byrow = TRUE
)

product_lv
```

Internally, `jsf_simulate()` calculates the net stoichiometry matrix as:

```text
nu = product - reactant
```

For this model, the net state changes are:

```text
Prey -> 2 Prey                 [+1,  0]
Predator -> ∅                  [ 0, -1]
Prey + Predator -> 2 Predator  [-1, +1]
```

### Run the simulation

Run the model and request a `JSFResult` object:

```{r}
result_lv <- jsf_simulate(
  x0 = c(prey = 50, predator = 10),
  rates = rates_lv,
  reactant = reactant_lv,
  product = product_lv,
  do_disc = c(1, 1),
  t_max = 10,
  dt = 0.01,
  switching_threshold = c(30, 30),
  species_names = c("prey", "predator"),
  return_type = "JSFResult"
)
```

The important arguments are:

* `x0` defines the initial number of prey and predators.
* `do_disc = c(1, 1)` allows both species to be treated discretely when required.
* `switching_threshold = c(30, 30)` defines the switching threshold for each species.
* `return_type = "JSFResult"` returns a structured result containing the simulation trajectory and associated metadata.

### Inspect the result

Print the result:

```{r}
result_lv
```

Calculate a summary:

```{r}
summary(result_lv)
```

Plot the simulated trajectories:

```{r}
plot(result_lv)
```

## SIR epidemic model

The SIR model describes the spread of an infectious disease through three compartments:

* `S`: susceptible;
* `I`: infectious;
* `R`: recovered.

The model contains two reactions:

1. Infection: `S + I -> 2 I`
2. Recovery: `I -> R`

The state vector is:

```{r}
x <- c(S, I, R)
```

### Define the model parameters

Set the total population, infection rate, and recovery rate:

```{r}
N <- 1000
beta <- 0.5
gamma <- 0.1
```

Here:

```text
N      = total population
beta   = infection rate
gamma  = recovery rate
```

### Define the rate function

Create a Python callable containing the infection and recovery rates:

```{r}
rates_sir <- reticulate::py_eval(sprintf(
  "lambda x, t: [%f * x[0] * x[1] / %f, %f * x[1]]",
  beta,
  N,
  gamma
))
```

The rates correspond to:

```text
beta * S * I / N    # infection
gamma * I           # recovery
```

### Define the stoichiometry

Define the reactant matrix:

```{r}
reactant_sir <- matrix(
  c(
    1, 1, 0,
    0, 1, 0
  ),
  ncol = 3,
  byrow = TRUE
)

reactant_sir
```

Define the product matrix:

```{r}
product_sir <- matrix(
  c(
    0, 2, 0,
    0, 0, 1
  ),
  ncol = 3,
  byrow = TRUE
)

product_sir
```

The net state changes are:

```text
S + I -> 2 I    [-1, +1,  0]
I -> R          [ 0, -1, +1]
```

### Run the simulation

```{r}
result_sir <- jsf_simulate(
  x0 = c(S = N - 10, I = 10, R = 0),
  rates = rates_sir,
  reactant = reactant_sir,
  product = product_sir,
  do_disc = c(0, 1, 0),
  t_max = 60,
  dt = 0.1,
  switching_threshold = c(N, 10, N),
  species_names = c("S", "I", "R"),
  return_type = "JSFResult"
)
```

In this example, `do_disc = c(0, 1, 0)` allows the infectious compartment to be treated discretely when required.

This is useful because stochastic effects can be particularly important when the number of infectious individuals is small.

### Inspect the result

Print the result:

```{r}
result_sir
```

Calculate a summary:

```{r}
summary(result_sir)
```

Plot the simulated trajectories:

```{r}
plot(result_sir)
```

## Return a data frame

The default output from `jsf_simulate()` is a regular R `data.frame`.

```{r}
df_lv <- jsf_simulate(
  x0 = c(prey = 50, predator = 10),
  rates = rates_lv,
  reactant = reactant_lv,
  product = product_lv,
  do_disc = c(1, 1),
  t_max = 10,
  dt = 0.01,
  switching_threshold = c(30, 30),
  species_names = c("prey", "predator"),
  return_type = "data.frame"
)

head(df_lv)
```

A data frame returned by `jsf_simulate()` contains:

* a `time` column; and
* one column for each model species.

A data frame can later be converted into a `JSFResult` object:

```{r}
result_from_df <- as_jsf_result(
  df_lv,
  method = "operator-splitting",
  config = list(
    model = "Lotka-Volterra",
    dt = 0.01,
    switching_threshold = c(30, 30)
  )
)

result_from_df
```

## Summary

The main `rjsf` workflow is:

```text
Define the initial state
  -> define the rate function
  -> define the reactant and product matrices
  -> call jsf_simulate()
  -> return a data.frame or JSFResult
  -> inspect the output with print(), summary(), and plot()
```

Use a `data.frame` when you primarily need the simulation values for further analysis.

Use a `JSFResult` when you want a structured object containing both the simulation trajectories and their associated metadata.
